The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-28 manuscript/manuscript.Rnw
move some tables and figures to supplemental material

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -1, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("scales")

@@

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and the standard Illumina protocol. Paired-end sequencing
was performed using the Illumina Genome Analyzer II with an insert
size of 230 base pairs (bp). Read lengths for species are shown in
\cref{tab:ref_genome_aligned_species}.

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \cref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \human or \mouse transcriptome
(\cref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \bonobo. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

\begin{figure}
<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
star.logs.unmapped <- 
    star.logs[grepl("%",field) & 
                  grepl("unmapped",field) & 
                      !grepl("too many mismatches",field),]
star.logs.unmapped <- 
    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
print(ggplot(star.logs.unmapped,
             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
      + geom_boxplot()
      + facet_grid(field~.,scales="free")
      + ylab("Percent of Reads")
      + xlab(NULL)
      + theme_bw()
      + theme(legend.position="top",
              plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
              axis.title.x=element_blank(),
              axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=1)
      )
pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
grid.text("A",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport()          
print(ggplot(star.logs[grepl("input reads$",field),],
             aes(x=capfirst(species),y=as.numeric(value)))
      + geom_boxplot()
      + scale_y_log10()
      + ylab("Reads")
      + xlab("Species")
      + theme_bw()
      + theme(legend.position="top",
              plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=2))
pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
grid.text("B",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport(2)
@ 
\caption{RNAseq read and alignment statistics for all species. A)
  Boxplot of percentages of unmapped reads in each species; 0\% of
  reads in all species were unmapped because of mismatches. B) Boxplot
  of number of RNAseq reads in each species.}
\label{fig:transcriptome_statistics}
\end{figure}

\Cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

\begin{table}
  \tiny
<<lau.lineage.specific,echo=FALSE,results="asis">>=
lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
setkey(lau.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in the Laurasiatheria
  superorder as compared to all other analyzed species.}
\label{tab:laurasiatheria_lineage_specific}
\end{table}

\begin{table}
  \tiny
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed between the
  Euasiatherian and Laurasiatherian superorder.}
\label{tab:laurasiatheria_eurasiatherian_lineage_specific}
\end{table}

We examined the expression of five Laurasiatherians: \cow, \sheep,
\pig, \horse, and \dog. We were unable to detect any significant
differences after correcting for multiple testing between
Laurasiatherians and all other analyzed species
(\cref{tab:laurasiatheria_lineage_specific}) or between
Laurasiatherians and Euasiatherians
(\cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \opossum are shown in
\cref{tab:top50_opossum_placenta} with a list of all genes in
Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
second highest expression is superoxide dismutase 3, extracellular
(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
which functions as an antioxidant. The gene with the second highest
expression is the hemoglobin subunit epsilon
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
which is expressed at the embryonic stage of development in
marsupials, as in other mammals
\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
subunits link up with protein products of two hemoglobin, zeta
(\textit{HBZ}) genes, which are also expressed to a high degree in the
opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
In addition, there is high expression of the gene hemoglobin, omega
(\textit{HBO},
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
% against Monodelphis domestica
a hemoglobin found only in marsupials and monotremes
\cite{Opazo.ea2008:Genomicevidenceforindependent}. The hemoglobin gene
\textit{HBA} is also expressed in the placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
Two cystatin genes
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
are expressed to a high degree, along with one \textit{CTSL} related
gene
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
\cref{fig:cathespin_evolution_tree}), which are known to interact with
cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
uncharacterized gene \textit{LOC10001195}
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
also shows high expression in the opossum placenta. The gene \textit{WFDC2}, a
member of the whey acidic protein (WAP) family is also expressed to a
high degree in the opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
Other genes with high expression include members of the annexin
complex
(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
In addition, there was little production of CYP19A1 and HSD3B1
(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
respectively).

\begin{figure}
<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of \textit{HBE1} and related genes}
\label{fig:hbe1_evolution_tree}
\end{figure}

@@

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]

cpt.distribution.shape <- 
    lapply(c("min_expression",
             "second_min_expression",
             "third_min_expression",
             "fourth_min_expression",
             "fifth_min_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
           })
cpt.distribution.shape <- 
    rbindlist(cpt.distribution.shape)

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Lowest Expression`:=
                               gsub("^min_expression$","First",
                                    `Lowest Expression`)]
cpt.distribution.shape <- 
    cpt.distribution.shape[,`Lowest Expression`:=
                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
                                           `Lowest Expression`,perl=TRUE),
                                      levels=c("First","Second","Third","Fourth","Fifth"))
                           ]
setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
      + geom_line(lwd=1)
      + xlab(expression(FPKM))
      + ylab("Genes with N-th Lowest Expression")
      + scale_y_log10()
      + scale_x_log10()
      + geom_vline(xintercept=10)
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=100,
                         color=`Lowest Expression`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Lowest Expression`],
                     arrow=arrow(length=unit(0.5,"cm"))
                   )
      + geom_text(mapping=aes(x=100,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Lowest Expression`],
                  show_guide=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title="Lowest\nExpression"))
      )
@ 
\caption{Distribution shape of placenta core genes. The vertical axis shows
  the number of orthologous gene sets included in the core placenta
  transcriptome if the least expression cutoff (or second-least, etc.)
  was set to the horizontal axis. For example, if the cutoff is set to 10
  (vertical black line), 37 genes would be included if the FPKM of all
  species must be at least 10, 134 genes would be included if all but
  one species must have FPKM of at least 10. 655 genes would be
  included if all but five species must have FPKM of at least 10.}
\label{fig:core_placenta_transcriptome_shape}
\end{figure}

\begin{table}
\footnotesize
<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
placenta.go.results <- placenta.go.results[,Annotated:=NULL]
setnames(placenta.go.results,"Significant","# Core")
setnames(placenta.go.results,"classic","p value")
setnames(placenta.go.results,"classic.fdr","FDR")
setnames(placenta.go.results,"GO.ID","GO ID")
setnames(placenta.go.results,"Term","GO Term")
print(xtable(placenta.go.results,
             digits=2,
             display=c("d","s","s","d","f","e","e","s")),
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE
      )
@ 
\caption{\ac{GO} terms most significantly enriched
  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]},
  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
  in the core placenta transcriptome as compared to all genes
  with 1:1 human orthologs in all species. “\# Core” is the
  number of genes in the core transcriptome for each \ac{GO} term;
  “Expected” is the number of genes expected in the core
  transcriptome for each term; “p-value” are the odds of this
  occurring by chance calculated by the Fisher exact test for
  each term; “FDR” is the false discovery rate as estimated by
  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}}
\label{tab:core_placenta_go_terms}
\end{table}

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
\Cref{tab:core_placenta_go_terms} lists the significant enrichment of
\ac{GO} terms in the core placenta transcriptome; notable terms
include those involved in cytoskeletal structure, protein transport,
membrane protein structure, and inflammation resolution.

@@

@ 
\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
  in 12 or more of the species analyzed from genes in the top 5\% of
  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.}
\label{tab:go_terms_all_species}
\end{table}

>>>>

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 0)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("scales")

@@

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and the standard Illumina protocol. Paired-end sequencing
was performed using the Illumina Genome Analyzer II with an insert
size of 230 base pairs (bp). Read lengths for species are shown in
Supplemental \cref{tab:ref_genome_aligned_species}.

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled 
using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
and annotated using Diamond
\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
\human or \mouse transcriptome
(Supplemental \cref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \bonobo. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

We examined the expression of five Laurasiatherians: \cow, \sheep,
\pig, \horse, and \dog. We were unable to detect any significant
differences after correcting for multiple testing between
Laurasiatherians and all other analyzed species
(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
Laurasiatherians and Euasiatherians
(Supplemental \cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \opossum are shown in
\cref{tab:top50_opossum_placenta} with a list of all genes in
Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
second highest expression is superoxide dismutase 3, extracellular
(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
which functions as an antioxidant. The gene with the second highest
expression is the hemoglobin subunit epsilon
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
which is expressed at the embryonic stage of development in
marsupials, as in other mammals
\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
subunits link up with protein products of two hemoglobin, zeta
(\textit{HBZ}) genes, which are also expressed to a high degree in the
opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
In addition, there is high expression of the gene hemoglobin, omega
(\textit{HBO},
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
% against Monodelphis domestica
a hemoglobin found only in marsupials and monotremes
\cite{Opazo.ea2008:Genomicevidenceforindependent} (Supplemental
\cref{fig:hbe1_evolution_tree}). The hemoglobin gene \textit{HBA} is
also expressed in the placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
Two cystatin genes
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
are expressed to a high degree, along with one \textit{CTSL} related
gene
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
\cref{fig:cathespin_evolution_tree}), which are known to interact with
cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
uncharacterized gene \textit{LOC10001195}
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
also shows high expression in the opossum placenta. The gene
\textit{WFDC2}, a member of the whey acidic protein (WAP) family is
also expressed to a high degree in the opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
Other genes with high expression include members of the annexin
complex
(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
In addition, there was little production of CYP19A1 and HSD3B1
(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
respectively).

@@

\begin{table}
\footnotesize
<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
placenta.go.results <- placenta.go.results[,Annotated:=NULL]
setnames(placenta.go.results,"Significant","# Core")
setnames(placenta.go.results,"classic","p value")
setnames(placenta.go.results,"classic.fdr","FDR")
setnames(placenta.go.results,"GO.ID","GO ID")
setnames(placenta.go.results,"Term","GO Term")
print(xtable(placenta.go.results,
             digits=2,
             display=c("d","s","s","d","f","e","e","s")),
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE
      )
@ 
\caption{\ac{GO} terms most significantly enriched
  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
  in the core placenta transcriptome as compared to all genes
  with 1:1 human orthologs in all species. “\# Core” is the
  number of genes in the core transcriptome for each \ac{GO} term;
  “Expected” is the number of genes expected in the core
  transcriptome for each term; “p-value” are the odds of this
  occurring by chance calculated by the Fisher exact test for
  each term; “FDR” is the false discovery rate as estimated by
  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}.}
\label{tab:core_placenta_go_terms}
\end{table}

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
\Cref{tab:core_placenta_go_terms} lists the significant enrichment of
\ac{GO} terms in the core placenta transcriptome; notable terms
include those involved in cytoskeletal structure, protein transport,
membrane protein structure, and inflammation resolution.

@@

@ 
\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
  in 12 or more of the species analyzed from genes in the top 1\% of
  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.}
\label{tab:go_terms_all_species}
\end{table}
